### 2015-05-20 Harris_2015/manuscript-materials/draft.tex
ready to submit, maybe?

Ecologists' intense interest in drawing inferences about species
interactions ---especially competition---from presence-absence data has
a long history (MacArthur 1958, Diamond 1975, Connor et al. 2013). If
nontrophic species interactions are important drivers of community
assembly patterns, then we should expect to see their influence in our
data sets. Despite decades of work and several major controversies,
however (Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009). More generally, it
can be difficult to reason about the complex web of direct and indirect
interactions that contribute to the structure of real assemblages,
especially when these interactions occur against a background of other
ecological processes such as dispersal and environmental filtering
(Connor et al. 2013). For this reason, it isn't always clear what kinds
of patterns would even constitute evidence of competition, as opposed to
some other biological process or random sampling error (Lewin 1983,
Roughgarden 1983).

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that
would be expected if \emph{all} species on the landscape were
independent (Strong et al. 1984, Gotelli and Ulrich 2009). Examining a
species pair against such a ``null'' background, however, rules out the
possibility that the overall association between two species could be
driven by an outside force, as shown in Figure 1. Even though the two
shrub species in this example compete with one another for resources at
a mechanistic level, they end up clustering together on the landscape
because they both grow best in areas that are not overshadowed by trees.
Due to fundamental constraints on the structure of valid covariance
matrices (Brown et al. 2010), this kind of positive correlation is
almost inevitable, even when all the interactions are negative. For this
reason, knowing that the two shrubs co-occur more than expected by
chance under some null model (e.g.~that their correlation in the data
set is positive) can not---by itself---tell us much about their direct
effects on one another.

While the competition between the two shrubs in the previous example
does not leave the commonly-expected pattern in community structure
(negative association at the landscape level), it nevertheless does
leave a signal in the data (Figure 1C). Specifically, \emph{among shaded
sites}, there will be a deficit of co-occurrences, and \emph{among
unshaded sites}, there will also be such a deficit. These
\emph{conditional} (``all-else-equal'') relationships can be estimated
with partial correlations (Albrecht and Gotelli 2001), although this
approach is best-suited for Gaussian variables. For non-Gaussian data, a
more general approach is needed.

In this paper, I introduce Markov networks (undirected graphical models
also known as Markov random fields; Murphy 2012) as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov
networks, which generalize partial correlations to non-Gaussian data,
have been used in many scientific fields to model associations between
various kinds of ``particles''. For example, a well-studied network
called the Ising model has played an important role in our understanding
of physics (where nearby particles tend to align magnetically with one
another; Cipra 1987). In spatial contexts, these models have been used
to describe interactions between adjacent grid cells (Harris 1974,
Gelfand et al. 2005). In neurobiology, they have helped researchers
determine which neurons are connected to one another by modeling the
structure in their firing patterns (Schneidman et al. 2006). Following
recent work by Azaele et al. (2010) and Fort (2013), I suggest that
ecologists could similarly treat species as the interacting particles in
this modeling framework. Doing so would allow ecologists to simulate and
study the landscape-level consequences of arbitrary species interaction
matrices, even when our observations are not Gaussian. While ecologists
explored some related approaches in the 1980's (Whittam and
Siegel-Causey 1981), computational limitations had previously forced
researchers to rely on severe approximations that produced
unintelligible results (e.g. ``probabilities'' greater than one; Gilpin
and Diamond 1982). Now that it is computationally feasible to fit these
models exactly, the approach has become worth a second look.

@@

\textbf{\emph{Conditional relationships and Markov networks.}}
Ecologists are often interested in inferring direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate
\emph{conditional} (``all-else-equal'') relationships, rather than
\emph{marginal} (``overall'') relationships. The most familiar
conditional relationship is the partial correlation, which indicates the
portion of the correlation between two species that remains after
controlling for other variables in the data set (Albrecht and Gotelli
2001), as opposed to the simple correlation, which does not control for
other variables. The example with the shrubs and trees in Figure 1 shows
how the two correlation measures can have opposite signs, and suggests
that the partial correlation is more relevant for drawing inferences
about species interactions (e.g.~competition). To the extent that our
observations are not multivariate Gaussian (e.g. presence-absence data),
partial covariances do not describe variables' conditional relationships
exactly, and a more general approach is needed. Markov networks allow
researchers to generalize conditional relationships to non-Gaussian
networks, much as generalized linear models allow researchers to extend
regression to non-Gaussian response variables (Lee and Hastie 2012).

@@

A major feature of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of
\(\beta\) coefficients (Murphy 2012). For example, if the coefficient
linking two mutualist species is \(+2\), then---all else equal---the
odds of observing either species increase by a factor of \(e^2\) when
its partner is present (Murphy 2012). Of course, if all else is
\emph{not} equal (e.g.~Figure 1, where the presence of one competitor is
associated with release from another competitor), then species' marginal
association rates can differ from this expectation. For this reason, it
is important to consider how coefficients' effects propagate through the
network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which
ensures that the probabilities of all possible assemblages that could be
produced by the model sum to one (bottom of Figure 2B). While the
analyses in this paper involve exact computations, this approach quickly
becomes infeasible as the number of species increases: with \(2^N\)
possible assemblages of \(N\) species the number of bookkeeping
operations required for exact inference quickly spirals exponentially
into the billions. Numerous techniques are available for working with
Markov networks that keep the computations tractable, either through
analytic approximations (Lee and Hastie 2012) or Monte Carlo sampling
(Salakhutdinov 2008), but these techniques are beyond the scope of this
paper.

@@

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability
distribution defined by the interaction coefficients in that figure.
Each of the methods described below (a Markov network, two
correlation-based methods and a null modeling approach) was then
evaluated on its ability to correctly infer that all three species
interactions were negative, despite the frequent co-occurrences of the
two shrub species.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500,
or 2500 observed communities (24 replicate simulations for each
combination). These simulated data sets span the range from small,
single-observer data sets to large collaborative efforts such as the
North American Breeding Bird Survey. As described in Appendix 1, I
randomly drew the ``true'' coefficient values for each replicate so that
most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.
I then used Gibbs sampling to randomly generate landscapes with varying
numbers of species and sites via Markov chain Monte Carlo (Appendix 1).
For half of the simulated landscapes, I treated each species' \(\alpha\)
coefficient as a constant, as described above. For the other half, I
treated the \(\alpha\) coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix 1). The latter set of simulated landscapes provide
an important test of the methods' ability to distinguish co-occurrence
patterns that were generated from pairwise interactions among the
observed species from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013, Blois et al. 2014).

\textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
presence-absence data.}} The previous sections involved known values of
\(\alpha\) and \(\beta\). In most real cases, however, the true values
of these parameters are not known, and ecologists will need to estimate
them from data. When the number of species is reasonably small, one can
compute exact maximum likelihood estimates for all of the \(\alpha\) and
\(\beta\) coefficients by optimizing Equation 1 (Appendix 2). Doing so
allows us to find the unique set of \(\alpha\) and \(\beta\)
coefficients that would be expected to produce exactly the observed
occurrence frequencies and co-occurrence frequencies. Fully-observed
Markov networks like the ones considered here have unimodal likelihood
surfaces (Murphy 2012), ensuring that this procedure will always
converge on the global maximum.

For the analyses in this paper, I made a small modification to this
maximum likelihood procedure. Given the large number of parameters
associated with some of the networks to be estimated, I regularized the
likelihood using a logistic prior distribution (Gelman et al. 2008) with
a scale of 1 on the \(\alpha\) and \(\beta\) terms.

The optimization procedures described here and in Appendix 2 are
implemented in the \emph{rosalia} package\footnote{source code available
  for download from https://github.com/davharris/rosalia} for the R
programming language (R Core Team 2015).

\textbf{\emph{Other inference techniques for comparison.}} After fitting
Markov networks to the simulated landscapes described above, I used
several other techniques for inferring the sign and strength of marginal
associations between pairs of species.

@@

The third method, described in Gotelli and Ulrich (2009), involved
simulating possible landscapes from a null model that retained the row
and column sums of the original matrix (Strong et al. 1984). Using the
default options in the Pairs software described in Gotelli and Ulrich
(2009), I simulated the null distribution of scaled C-scores (a test
statistic describing the number of \emph{non}-co-occurrences between two
species). The software then calculated a \(Z\)-statistic for each
species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

@@

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate\footnote{The null model generated one \(Z\)-score
  outlier greater than 1000, which dominated the regression and squared
  error analyses. To reduce its influence on these results, I changed
  its value to 32.5, which was the value of the next largest \(Z\)-score
  in the null model's results.}. Then, for each combination of species
richness, landscape size, method, and presence/absence of environmental
filtering, I calculated the mean squared error associated with the
rescaled estimate across all 12 replicates. Finally, I calculated an
\(R^2\) value for each landscape by comparing this squared error
estimate with the squared error associated with the null hypothesis that
all interactions were zero.

@@

\textbf{\emph{Three species.}} As shown in Figure 1, the marginal
relationship between the two shrub species was positive---despite their
competition for space at a mechanistic level--- due to indirect effects
of the dominant tree species. As a result, the covariance method falsely
reported positive associations 94\% of the time and the
randomization-based null model falsely reported such associations 100\%
of the time. The two methods for evaluating conditional relationships
(Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each
correctly identified the direct negative interaction between the shrubs
92\% of the time.

\textbf{\emph{Larger landscapes.}} The Markov network consistently
outperformed all of the other techniques that I evaluated (Figure 3). Of
the remaining methods, partial covariances tended to be most accurate,
especially when 100 or more communities had been observed on the
landscape. \(Z\)-scores from the null model had the least predictive
utility overall, as indicated by \(R^2\) values. In general, all the
models' accuracies declined when environmental filters were added, but
the declines were smaller when environmentally-induced correlations were
spread out among a larger number of pairwise interactions.

@@

The results presented above are very promising, as they show that Markov
networks can recover much of the variation in species' pairwise
interaction strengths from observational data, even when direct
interactions are largely overwhelmed by indirect effects (e.g.~Figure 1)
or environmental effects (lower panels of Figure 3). For cases where it
is infeasible to fit a Markov network, these results also indicate that
partial covariances---which can be computed straightforwardly by linear
regression---can often provide an accurate approximation.

Apart from the environmental filters, the simulated landscapes presented
here represent the best-case scenario for these methods. Future research
should thus examine these models' performance characteristics when the
``true'' interaction matrices include guild structure or trophic levels,
which could make the \(\beta\) coefficients much more difficult to infer
(particularly for linear approximations like the partial covariance
approach; Loh and Wainwright (2013)). On the other hand, ecologists may
often have prior information about the nature of real species'
interaction patterns from natural history or ecological experiments,
which could substantially reduce the probability and magnitude of error.
The rosalia package has built-in mechanisms for incorporating this kind
of information, if it can be expressed as a prior probability
distribution or a penalty on the likelihood.

@@

Despite these limitations, the results with environmental filtering seem
to indicate that the method can be very robust. Additionally, the fact
that Markov networks provide a likelihood function to optimize makes
them highly extensible, even when it is inconvenient to compute the
likelihood exactly. For example, the mistnet software package for joint
species distribution modeling (Harris 2015) can fit \emph{approximate}
Markov networks to large species assemblages (\textgreater{}100 species)
while simultaneously modeling each species' response to the abiotic
environment with complex, nonlinear functions. This sort of approach,
which combines multiple ecological processes, could help ecologists to
disentangle different factors behind the co-occurrence patterns we
observe in nature. Numerous other extensions are possible: similar
networks can be fit with continuous variables, count data, or both (Lee
and Hastie 2012). There are even methods (Whittam and Siegel-Causey
1981, Tjelmeland and Besag 1998) that would allow the coefficient
linking two species in an interaction matrix to vary as a function of
the abiotic environment or of third-party species that could tip the
balance between facilitation and exploitation (Bruno et al. 2003). Fully
exploring these possibilities will require more research into the
various available approximations to the log-likelihood and to its
gradient, in order to balance efficiency, accuracy, and the ability to
generate confidence limits for statistical inference.

@@

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for
nearly a century, and the models' properties, capabilities, and limits
are well-understood in a huge range of applications, from spatial
lattices (Gelfand et al. 2005) to haphazard neural connections in the
retina and brain (Schneidman et al. 2006) to complex behavioral
interactions among human beings (Lee et al. 2013). Modeling species
interactions using the same framework would thus allow ecologists to tap
into an enormous set of existing discoveries and techniques for dealing
with indirect effects, stability, and alternative stable states.

Finally, the null models commonly used by ecologists for drawing
inferences about pairwise species interactions do not seem to
distinguish between direct and indirect interactions any better than
simple correlation coefficients. Null and neutral models can be very
useful for clarifying our thinking about the numerical consequences of
species' richness and abundance patterns (Harris et al. 2011, Xiao et
al. 2015), but it is less clear whether deviations from the null
hypothesis can clearly implicate specific biological processes like
competition.

@@

\textbf{\emph{Figure 3.}} Proportion of variance in interaction
coefficients explained by each method with 5, 10, or 20 species arrayed
across varying numbers of sampled locations when environmental filtering
was absent (top row) or present (bottom row).

@@

Brown, I. H., E. Bedrick, S. Morgan, I.-L. E. C. Ernest, and I. F.
Kelly. 2010. Constraints on Negative Relationships: Mathematical Causes
and Ecological. The Nature of Scientific Evidence: Statistical,
Philosophical, and Empirical Considerations:298.

>>>>

Ecologists' intense interest in drawing inferences about species
interactions ---especially competition---from presence-absence data has
a long history (MacArthur 1958, Diamond 1975, Connor et al. 2013). If
nontrophic species interactions are important drivers of community
assembly patterns, then we should expect to see their influence in our
data sets. Despite decades of work and several major controversies,
however (Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009). More generally, it
can be difficult to reason about the complex web of direct and indirect
interactions in real assemblages, especially when these interactions
occur against a background of other ecological processes such as
dispersal and environmental filtering (Connor et al. 2013). For this
reason, it isn't always clear what kinds of patterns would even
constitute evidence of competition, as opposed to some other biological
process or random sampling error (Lewin 1983, Roughgarden 1983).

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that
would be expected if \emph{all} species on the landscape were
independent (Strong et al. 1984, Gotelli and Ulrich 2009). Examining a
species pair against such a ``null'' background, however, rules out the
possibility that the overall association between two species could be
driven by an outside force. For example, even though the two shrub
species in Figure 1 compete with one another for resources at a
mechanistic level, they end up clustering together on the landscape
because they both grow best in areas that are not overshadowed by trees.
If this sort of effect is common, then significant deviations from
independence will not---by themselves---provide convincing evidence of
species' direct effects on one another.

While the competition between the two shrubs in the previous example
does not leave the commonly-expected pattern in community structure
(negative association at the landscape level), it nevertheless does
leave a signal in the data (Figure 1C). Specifically, \emph{among shaded
sites}, there will be a deficit of co-occurrences, and \emph{among
unshaded sites}, there will also be such a deficit.

In this paper, I introduce Markov networks (undirected graphical models
also known as Markov random fields; Murphy 2012) as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them with observational data. Markov
networks, which generalize partial correlations to non-Gaussian data,
have been used in many scientific fields to model associations between
various kinds of ``particles''. For example, a well-studied network
called the Ising model has played an important role in our understanding
of physics (where nearby particles tend to align magnetically with one
another; Cipra 1987). In spatial contexts, these models have been used
to describe interactions between adjacent grid cells (Harris 1974,
Gelfand et al. 2005). In neurobiology, they have helped researchers
determine which neurons are connected to one another by modeling the
structure in their firing patterns (Schneidman et al. 2006). Following
recent work by Azaele et al. (2010) and Fort (2013), I suggest that
ecologists could similarly treat species as the interacting particles in
this modeling framework. Doing so would allow ecologists to simulate and
study the landscape-level consequences of arbitrary species interaction
matrices, even when our observations are not Gaussian. While ecologists
explored some related approaches in the 1980's (Whittam and
Siegel-Causey 1981), computational limitations had previously imposed
severe approximations that produced unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982). Now that
it is computationally feasible to fit these models exactly, the approach
has become worth a second look.

@@

\textbf{\emph{Conditional relationships and Markov networks.}}
Ecologists are often interested in inferring direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate
\emph{conditional} (``all-else-equal'') relationships, rather than
\emph{marginal} (``overall'') relationships. The most familiar
conditional relationship is the partial correlation, which indicates the
portion of the sample correlation between two species that remains after
controlling for other variables in the data set (Albrecht and Gotelli
2001). The example with the shrubs and trees in Figure 1 shows how the
two correlation measures can have opposite signs, and suggests that the
partial correlation is more relevant for drawing inferences about
species interactions (e.g.~competition). Markov networks extend this
approach to non-Gaussian data, much as generalized linear models do for
linear regression (Lee and Hastie 2012).

@@

A major benefit of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of
\(\beta\) coefficients (Murphy 2012). For example, if the coefficient
linking two mutualist species is \(+2\), then---all else equal---the
odds of observing either species increase by a factor of \(e^2\) when
its partner is present (Murphy 2012). Of course, if all else is
\emph{not} equal (e.g.~Figure 1, where the presence of one competitor is
associated with release from another competitor), then species' marginal
association rates can differ from this expectation. For this reason, it
is important to consider how coefficients' effects propagate through the
network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which
ensures that the probabilities of all possible assemblages that could be
produced by the model sum to one (bottom of Figure 2B). Calculating
\(Z(\alpha, \beta)\) exactly, as is done in this paper, quickly becomes
infeasible as the number of species increases: with \(2^N\) possible
assemblages of \(N\) species the number of bookkeeping operations
required for exact inference quickly spirals exponentially into the
billions. Numerous techniques are available for working with Markov
networks that keep the computations tractable, either through analytic
approximations (Lee and Hastie 2012) or Monte Carlo sampling
(Salakhutdinov 2008), but they are beyond the scope of this paper.

@@

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability
distribution defined by the interaction coefficients in that figure
(Appendix A). Each of the methods described below (a Markov network, two
correlation-based methods and a null modeling approach) was then
evaluated on its ability to correctly infer that the two shrub species
competed with one another, despite their frequent co-occurrence.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500,
or 2500 observed communities (24 replicate simulations for each
combination; Appendix B). These simulated data sets span the range from
small, single-observer data sets to large collaborative efforts such as
the North American Breeding Bird Survey. As described in Appendix B, I
randomly drew the ``true'' coefficient values for each replicate so that
most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.
I then used Gibbs sampling to randomly generate landscapes with varying
numbers of species and sites via Markov chain Monte Carlo (Appendix C).
For half of the simulated landscapes, I treated each species' \(\alpha\)
coefficient as a constant, as described above. For the other half, I
treated the \(\alpha\) coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix B). The latter set of simulated landscapes provide
an important test of the methods' ability to distinguish co-occurrence
patterns that were generated from pairwise interactions among the
observed species from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013, Blois et al. 2014).

\textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
presence-absence data.}} The previous sections involved known values of
\(\alpha\) and \(\beta\). In practice, ecologists will often need to
estimate these parameters from data instead. When the number of species
is reasonably small, one can compute exact maximum likelihood estimates
for all of the \(\alpha\) and \(\beta\) coefficients by optimizing
Equation 1 (Appendix C). Doing so allows us to find the unique set of
\(\alpha\) and \(\beta\) coefficients that would be expected to produce
exactly the observed co-occurrence frequencies. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
(Murphy 2012), ensuring that this procedure will always converge on the
global maximum. The optimization procedures described here and in
Appendix C are implemented in the \emph{rosalia} package
({\textbf{???}}) for the R programming language (R Core Team 2015).

For analyses with 5 or more species, I made a small modification to the
maximum likelihood procedure described above. Given the large number of
parameters associated with some of the networks to be estimated, I
regularized the likelihood using a logistic prior distribution (Gelman
et al. 2008) with a scale of 1 on the \(\alpha\) and \(\beta\) terms.

\textbf{\emph{Other inference techniques for comparison.}} After fitting
Markov networks to the simulated landscapes described above, I used
several other techniques for inferring the sign and strength of marginal
associations between pairs of species (Appendix B).

@@

The third method, described in Gotelli and Ulrich (2009), involved
simulating possible landscapes from a null model that retains the row
and column sums of the original matrix (Strong et al. 1984). Using the
default options in the Pairs software described in Gotelli and Ulrich
(2009), I simulated the null distribution of scaled C-scores (a test
statistic describing the number of \emph{non}-co-occurrences between two
species). The software then calculated a \(Z\)-statistic for each
species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

@@

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate\footnote{The null model generated one \(Z\)-score
  outlier greater than 1000, which dominated the regression and squared
  error analyses. To reduce its influence on these results, I changed
  its value to 32.5, which was the value of the next largest \(Z\)-score
  in the null model's results.}. For each estimate of a species pair's
interactions, I used this regression to calculate the squared error
associated with method that produced it. Finally, I averaged these
squared errors for each combination of species richness, landscape size,
statistical method, and presence/absence of environmental filtering
across all 12 replicates; the mean squared errors associated with these
subsets of the data determined the proportion of variance explained by
each method under different conditions.

@@

\textbf{\emph{Three species.}} As shown in Figure 1, the marginal
relationship between the two shrub species was positive---despite their
competition for space at a mechanistic level--- due to indirect effects
of the dominant tree species. As a result, the covariance method falsely
reported positive associations 94\% of the time and the
randomization-based null model falsely reported such associations 100\%
of the time. The two methods for evaluating conditional relationships
(Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each
correctly identified the direct negative interaction between the shrubs
94\% of the time.

\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
methods varied substantially, depending on the parameters that produced
the simulated communities (Figure 3). In general, however, there was a
consistent ordering: the Markov network explained 54\% of the variance
overall, followed by partial covariances (32\%), sample covariances
(21\%), and \(Z\) scores from the null model (17\%).\\The models'
accuracies tended to decline when environmental filters were added,
particularly when the number of species was small and the effects could
not be diluted among many pairwise interactions.

@@

The results presented above are very promising, as they show that Markov
networks can recover species' pairwise interactions from observational
data, even when direct interactions are largely overwhelmed by indirect
effects (e.g.~Figure 1) or environmental effects (lower panels of Figure
3). For cases where it is infeasible to fit a Markov network, these
results also indicate that partial covariances---which can be computed
straightforwardly by linear regression---can often provide an accurate
approximation.

Apart from the environmental filters, the simulated landscapes presented
here represent the best-case scenario for these methods. Future research
should thus examine these models' performance characteristics when the
``true'' interaction matrices include guild structure or trophic levels,
which could make the \(\beta\) coefficients much more difficult to infer
(particularly for linear approximations like the partial covariance
approach; Loh and Wainwright (2013)). On the other hand, ecologists may
often have prior information about the nature of real species'
interaction patterns from natural history or ecological experiments,
which could substantially reduce the probability and magnitude of error.
The rosalia package ({\textbf{???}}) has built-in mechanisms for
incorporating this kind of information, if it can be expressed as a
prior probability distribution or a penalty on the likelihood.

@@

Despite these limitations, the results with environmental filtering seem
to indicate that the method can be very robust. Additionally, the fact
that Markov networks provide a likelihood function to optimize makes
them highly extensible, even when it is inconvenient to compute the
likelihood exactly. For example, the mistnet software package for joint
species distribution modeling (Harris 2015) can fit \emph{approximate}
Markov networks to large species assemblages (\textgreater{}300 species)
while simultaneously modeling each species' response to the abiotic
environment with complex, nonlinear functions. This sort of approach,
which combines multiple ecological processes, could help ecologists to
disentangle different factors behind the co-occurrence patterns we
observe in nature. Numerous other extensions are possible: similar
networks can be fit with continuous variables, count data, or both (Lee
and Hastie 2012). There are even methods (Whittam and Siegel-Causey
1981, Tjelmeland and Besag 1998) that would allow the coefficient
linking two species in an interaction matrix to vary as a function of
the abiotic environment or of third-party species that could tip the
balance between facilitation and exploitation (Bruno et al. 2003). Fully
exploring these possibilities will require more research into the
various available approximations to the log-likelihood and to its
gradient, in order to balance efficiency, accuracy, and the ability to
generate confidence limits for statistical inference.

@@

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for
nearly a century, and the models' properties, capabilities, and limits
are well- understood in a huge range of applications, from spatial
modeling (Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006)
to models of human behavior {[}Lee et al. (2013);
salakhutdinov\_restricted\_2007{]}. Modeling species interactions using
the same framework would thus allow ecologists to tap into an enormous
set of existing discoveries and techniques for dealing with indirect
effects, stability, and alternative stable states.

These results also have important implications for the continued use of
fixed-fixed null models in ecology. The small simulated landscapes
described by Figure 1 show that test statistics based on marginal
co-occurrence (such as C-scores) will not always have a straightforward
relationship with the underlying ecological processes. Moreover, the
larger communities analyzed in Figure 3 often fell so far outside the
null distribution\footnote{Nearly 20\% of the species pairs fell outside
  the 99.99994\% confidence intervals implied by their \(Z\) scores
  (\(|Z| > 5\)), and about 10\% had uncorrected p-values below R's
  default numerical precision of \(2 \times 10^{-16}\).} that it
probably makes more sense to reject whole model rather than to assign
blame for the discrepancy to any one species pair. On average, the
pairwise \(Z\) scores from the null model provided less information
about direct species interactions than correlation coefficients did.
Researchers using the null modeling approach may be able to predict
twice as much of the variance in species' ``true'' interaction strengths
using partial covariances from linear regression or triple them using a
Markov network.

Null and neutral models can be very useful for clarifying our thinking
about the numerical consequences of species' richness and abundance
patterns (Harris et al. 2011, Xiao et al. 2015), but deviations from a
null model must be interpreted with care (Roughgarden 1983). In complex
networks of ecological interactions---and even in small networks with
three species---it may simply not be possible to implicate individual
species pairs or specific ecological processes like competition by
rejecting a general- purpose null. Direct estimates of species'
conditional associations may be the only way to make these inferences
reliably.

@@

\textbf{\emph{Figure 3.}} Proportion of variance in interaction
coefficients explained by each method with 5, 10, or 20 species arrayed
across varying numbers of sampled locations when environmental filtering
was absent (top row) or present (bottom row). A negative \(R^2\) values
implies that the squared error associated with the corresponding subset
of the predictions was larger than the error one would get from assuming
that all coefficients equalled zero.

@@

### 2015-05-20 Harris_2015/models.Rmd
ready to submit, maybe?

```{r, cache = TRUE}
rosalia_results = fit_model("rosalia", grep("^[^2]", dir("fakedata/matrices"), value = TRUE))
```

@@

results = do.call(
  rbind,
  lapply(
    list(
      rosalia_results, 
      cov_results, 
      partial_results, 
      pairs_results
    ),
    function(x){
      residuals = residuals(
        lm(
          truth ~ predicted + 0, 
          data = x, 
          weights = 1/choose(n_spp, 2),
          y = TRUE
        )
      )
      cbind(x, residuals = residuals)
    }
  )
)

@@

>>>>

```{r, cache = TRUE}
rosalia_results = fit_model("rosalia", dir("fakedata/matrices"))
```

@@

results = do.call(
  rbind,
  lapply(
    list(
      rosalia_results, 
      cov_results, 
      partial_results, 
      pairs_results
    ),
    function(x){
      linear_model = lm(
        truth ~ predicted + 0, 
        data = x, 
        weights = 1/choose(n_spp, 2),
        y = TRUE
      )
      cat(x$method[[1]])
      print(summary(linear_model))
      cat("\n", "\n", "\n")
      residuals = residuals(linear_model)
      cbind(x, residuals = residuals)
    }
  )
)

@@

What percentage of absolute $Z$ scores exceed 5?

```{r}
mean(abs(pairs_results$predicted) > 5)
```

What percentage of absolute $Z$ scores produce p-values below `.Machine$double.eps`?

```{r}
.Machine$double.eps

p_values = pnorm(1 - abs(pairs_results$predicted))
mean(p_values < .Machine$double.eps)
```
